---
title: "RNA editing during viral infection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{chikv-editing}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is an analysis of RNA editing during a Chikungunya (CHIKV)
infection of mouse lymph node tissue, described in:

Carpentier KS, Sheridan RM, Lucas CJ, Davenport BJ, Li FS, Lucas ED,
McCarthy MK, Reynoso GV, May NA, Tamburini BAJ, Hesselberth JR, Hickman
HD, Morrison TE. MARCO+ lymphatic endothelial cells sequester
arthritogenic alphaviruses to limit viremia and viral dissemination.
EMBO J. 2021 Nov 15;40(22):e108966. doi: 10.15252/embj.2021108966. Epub
2021 Oct 7. PMID: 34618370; PMCID: PMC8591538.

Mice were infected with an attenuated strain of CHIKV and lymph nodes
were harvested after 24 hours and analyzed by 10x Genomics 3'
single-cell RNA sequencing, focusing on CD45- stromal cells.

Data from triplicate infection and mock samples were analyzed by the
CellRanger pipeline, and output BAM files serve as the input data below.

-   Add quality control metrics here (cell numbers, levels of viral
    RNA?)

```{r libs, message=FALSE}
library(raer)
library(tidyverse)
library(cowplot)
library(fs)
library(qs)

library(BiocParallel)
library(SummarizedExperiment)
library(plyranges)
library(rtracklayer)
library(GenomicFeatures)
```

```{r setup_common_data}
ext_path <- fs::path('~/projects/raer-chikv')
samples <- c('A1','A2','A3', 'M1', 'M2', 'M3')

fns = list(
  bams    = fs::path(ext_path, 'bam', str_c( samples, '.bam')),
  bais    = fs::path(ext_path, 'bam', str_c(samples, '.bam.bai')),
  fasta   = fs::path(ext_path, 'ref', 'mm10_CHIKV_AF15561.fa.gz'),
  rmsk_db = fs::path(ext_path, 'ref', 'rmsk.mm10.tsv.gz'),
  snp_db  = fs::path(ext_path, 'ref', 'snp142.mm10.bed.gz'),
  gtf     = fs::path(ext_path, 'ref', 'gencode.knownGene.mm10.gtf.gz')
)
```

This is the main raer pileup routeine. On this data set, takes about 4
hours (parallelized) and generates a large RangedSummarizedExperiment
with \~48 million sites.

```{r run_pileup, eval=FALSE}
fp <- FilterParam(
  # min depth of 3 for each replicate
  min_nucleotide_depth = 3,
  min_base_quality = 30,
  min_mapq = rep(255, 6),
  library_type = rep("fr-second-strand", 6),
  trim_5p = 5,
  trim_3p = 5,
  indel_dist = 4,
  homopolymer_len = 6,
  max_mismatch_type = c(3, 3),
  min_read_bqual = c(0.25, 20),
  only_keep_variants = rep(TRUE, 6)
)

bpp <- SnowParam(workers = 24)
plps <- get_pileup(
  fns$bams,
  fafile = fns$fasta,
  filterParam = fp,
  BPPARAM = bpp
)

names(plps) <- samples 
plps

all_rse <- create_se(plps)

# remove multi-allelics now, they're not much use later.
# this step removes about 15% (~8 million) of the sites.
all_rse <- remove_multiallelic(all_rse)

chikv_rse <- subsetByOverlaps(
  all_rse,
  filter(rowRanges(all_rse), seqnames == 'CHIKV_AF15561')
)

mm_rse <- subsetByOverlaps(
  all_rse,
  filter(rowRanges(all_rse), seqnames != 'CHIKV_AF15561')
)

qs::qsave(all_rse, fs::path(ext_path, '/rds/all_rse.rds'), nthreads = 10)
qs::qsave(mm_rse, fs::path(ext_path, '/rds/mm_rse.rds'), nthreads = 10)
qs::qsave(chikv_rse, fs::path(ext_path, '/rds/chikv_rse.rds'), nthreads = 10)
```

If rds files have already been created, you can start from here.

```{r load_rds, eval = FALSE}
all_rse <- qs::qread(fs::path(ext_path, '/rds/all_rse.rds'), nthreads = 10)
mm_rse <- qs::qread(fs::path(ext_path, '/rds/mm_rse.rds'), nthreads = 10)
chikv_rse <- qs::qread(fs::path(ext_path, '/rds/chikv_rse.rds'), nthreads = 10)
```

## Filtering for high confidence sites

From the intial set of \~50 million sites from the pileup, we filter
based on several criteria:

1.  filter on `nVar`, requiring at least 3 reads for every sample among
    replicates

2.  remove clustered variants and variants near splice sites, as these
    tend to be lower confidence due to mapping artifacts

3.  after `calc_edit_frequency()`, filter on `depth` and `edit_freq`,
    requiring a minimum of 3 reads and an editing frequency of at least
    5%.

4.  annotate the remaining variants based on overlap with reference
    databases of repeats and SNPs, and remove variants based on ad hoc
    criteria

### Filtering sites by treshold

This is a helper function to filter samples within each replicate.

```{r filter_fxn}
filter_replicates <- function(rse, assay_type = NULL, thresh = 0, n_rep = 3) {
 
  x <- assay(rse, assay_type)[, 1:6]
  
  mocks <- c('M1', 'M2', 'M3')
  virus <- c('A1', 'A2', 'A3')

  keep_mocks <- rowSums(x[, mocks] >= thresh) >= n_rep
  keep_virus <- rowSums(x[, virus] >= thresh) >= n_rep 

  # keep anything that passes the filter in either mock or virus samples
  keep_all <- keep_mocks | keep_virus
  
  rse[keep_all, ]
}
```

The order of filtering here is important: the first `nVar` filter
removes a large chunk of the data, so we only compute the next steps on
\~10% of the sites.

```{r filter_mm}
# filter for at least 3 reads at each site every sample in the replicates.
# this removes a large chunk of the data that is low coverage, ~90% or so
# n = 861939 
mm_rse_filt <- filter_replicates(mm_rse, "nVar")

# Next, remove sites based on gene annotations
tx_db <- GenomicFeatures::makeTxDbFromGRanges(
  rtracklayer::import(fns$gtf)
)

# n = 325352
mm_rse_filt <- remove_clustered_variants(
  mm_rse_filt, tx_db, variant_dist = 100
  ) |>
  remove_splice_variants(tx_db)

# calculate `edit_freq` and `depth`
mm_rse_filt <- calc_edit_frequency(mm_rse_filt)

# filter on site depth >= 3 for all samples within replicates 
mm_rse_filt <- filter_replicates(mm_rse_filt, "depth", thresh = 3)

# filter for edit_freq >= 0.05 for at least 3 samples
# n = 266645
mm_rse_filt <- filter_replicates(mm_rse_filt, "edit_freq", thresh = 0.05)

mm_rse_filt
```

Now do the same for viral data --- these may be a bit harsh for the
virus coverage we have.

```{r filter_chikv}
# n = 4.25
chikv_rse_filt <- filter_replicates(chikv_rse, "nVar")

# calculate edit frequencies and depth
chikv_rse_filt <- calc_edit_frequency(chikv_rse_filt)

# filter on site depth >= 3 for all samples within replicates 
chikv_rse_filt <- filter_replicates(chikv_rse_filt, "depth", thresh = 3)

# filter for edit_freq >= 0.05 for all replicates
# n = 42
chikv_rse_filt <- filter_replicates(chikv_rse_filt, "edit_freq", thresh = 0.05)

chikv_rse_filt
```

### Filtering by annotation

Filter the mouse data by annotation, comparing to reference SNP and
repeat databases.

```{r annotate_rse}
# load external annotations
rmsk_db <-  read_tsv(
  fns$rmsk_db,
  col_names = c('seqnames', 'start', 'end', 'strand', 'repName', 'repClass', 'repFamily')
) |>
  plyranges::as_granges()

snp_db  <- read_tsv(
  fns$snp_db,
  col_names = c('seqnames', 'start', 'end', 'snpId', 'score', 'strand')
) |>
  plyranges::as_granges()

# annotate the filtered sites
mm_rse_filt <- annot_from_gr(
  mm_rse_filt,
  rmsk_db,
  cols_to_map = c("repName", "repFamily")
)

mm_rse_filt <- annot_from_gr(
  mm_rse_filt,
  snp_db,
  cols_to_map = "snpId"
)

# filter out specific repeat classes
# n = 244246
mm_rse_filt <- mm_rse_filt[!rowData(mm_rse_filt)$repFamily %in% c("Simple_repeat", "Low_Complexity")]

# filter out common variants (i.e., already have a snpId)
# n = 226446
mm_rse_filt <- mm_rse_filt[is.na(rowData(mm_rse_filt)$snpId)] 

mm_rse_filt
```

## Save filtered data

```{r save_filt_data, eval=FALSE}
qs::qsave(mm_rse_filt, fs::path(ext_path, '/rds/mm_rse_filt.rds'), nthreads = 10)
qs::qsave(chikv_rse_filt, fs::path(ext_path, '/rds/chikv_rse_filt.rds'), nthreads = 10)

mm_rse_filt <- qs::qread(fs::path(ext_path, '/rds/mm_rse_filt.rds'), nthreads = 10)
chikv_rse_filt <- qs::qread(fs::path(ext_path, '/rds/chikv_rse_filt.rds'), nthreads = 10)
```

## Exploratory analysis

### Tidy and examine the variants

```{r tidy_data}
tidy_mm_filt_vars <- assay(mm_rse_filt, "Var") |>
  as.data.frame() |>
  rownames_to_column("site") |>
  as_tibble() |>
  pivot_longer(-site) |>
  # filter out ambiguous Vars
  filter(value != '-' & !str_detect(value, '^N')) |>
  # polish columns in variables
  separate(name, into = c('treat', 'rep'), sep = 1) |>
  mutate(treat = case_when(
    treat == 'A' ~ 'virus',
    treat == 'M' ~ 'mock'
  )) |>
  separate(site, into = c('chrom', 'pos', 'strand'), sep = '_')
```

```{r count_and_plot}
# inspect the counts
filt_var_counts <- tidy_mm_filt_vars |>
  group_by(treat, rep) |>
  dplyr::count(value) |>
  # make variants more intutive, i.e. `AT` -> `A>T`
  separate(value, into = c('ref', 'alt'), sep = 1) |>
  unite(var, ref:alt, sep = '>') 

ggplot(filt_var_counts, aes(var, n, fill = rep)) +
  geom_col() +
  facet_grid(~treat) +
  scale_fill_brewer(palette = 'Reds') +
  theme_cowplot() +
  labs(
    x = 'variant',
    y = 'count'
  ) + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

The plot of counts above shows us a few things:

-   The most abundant variants are anything to A (`C>A`, `G>A`, `T>A`).
    I'm honestly not sure where that signal comes from, and Kent is
    stumped too.

    I considered calculating a normalized frequency comparing to the raw data,
    but (a) that's an expensive calculation and (b) most of those raw sites are
    low coverage anyway (see the `nVar` step at the beginning), so are going to
    be largely uninformative.

    I think the thing to do is to figure out how the remaining filters
    behave for that subset of sites. It's possible these `X>A` sites are
    generally lower coverage relative to the signal (`A>G`), and so
    improved thresholds for the filtering steps may eliminate them.

-   The next most abundant variant is the signal of adenosine editing
    and the one we're after (`A>G`). so all good there. There are about
    1e4 `A>G` sites per replicate:

        # A tibble: 6 × 4
        # Groups:   treat, rep [6]
        treat rep   value     n
        <chr> <chr> <chr> <int>
        1 A     1     AG    12323
        2 A     2     AG    12192
        3 A     3     AG    12289
        4 M     1     AG    11923
        5 M     2     AG    12236
        6 M     3     AG    12143

-   There's a small signal for `T>C` (not sure what that is), and the
    remaining are at much lower levels.


